---
title: "PERMANOVA on understory plant data"
author: "Ekaphan Kraichak"
date: '2022-11-01'
output: github_document
---

## load the libraries

```{r}
library(tidyverse)
library(vegan)
```

## import data

```{r}
env <- read.csv("Environmental_matrix.csv")
comm <- read.csv("UnderGrowth_matrix.csv")
```

## combine env and comm to select only the plot that experience fire to make sure that they are in the same order. Filter only the plot with fire
```{r}
comm.env <- env %>% 
  select(Plot, Fire, Fire.appear, forest.type) %>%
  mutate(forest.type = recode(forest.type, "MDF,bamboo" = "MDFB", "OpenArea" = "OA")) %>% 
  inner_join(comm, by = c("Plot", "Fire")) %>% 
  filter(Fire.appear == "Fire") %>% 
  filter(!(Plot %in% c(123, 124, 115))) ## these plots are OA and way outliers. Plot 115 is also acting weird

comm.env.E0 <- comm.env[,1:4] ## environmental matrix
comm.env.C0 <- comm.env[,5:484] 
```

after all the filtering, some species disappear from the matrix. We will select only species that occur in more than 20 plots (remove rare species)

```{r}
main.species <- names(which(colSums(comm.env.C0 > 0) > 20))


comm.env.C1 <- comm.env.C0[,main.species]
## find not empty row
not.empty <- which(rowSums(comm.env.C1) > 0)

comm.env.C <- wisconsin(comm.env.C1[not.empty, main.species])
## community matrix with Wisconsin double standardization
comm.env.E <- comm.env.E0[not.empty, ]

```


## run PERMANOVA using adonis

```{r}
perm <- how(nperm = 1999)
setBlocks(perm) <- with(comm.env.E, forest.type)
adonis2(comm.env.C ~ Fire*forest.type, data = comm.env.E, permutations = perm)
```

```{r}
xx <- betadisper(vegdist(comm.env.C), comm.env.E$Fire)
anova(xx)
```

## visualize the data

```{r, fig.height=5, fig.width=8}
mod <- metaMDS(comm.env.C)

mod_scores <- scores(mod)

## without hull
mod_score_env <- cbind(comm.env.E, mod_scores) %>% 
  mutate(forest = paste(forest.type, Fire, sep = " ")) %>% 
  mutate(forest = factor(forest, levels = c("DDF before", "DDF after", "MDF before", "MDF after", "MDFB before", "MDFB after"))) %>% 
  mutate(Fire = factor(Fire, levels = c("before","after")))

## with hull
mod_score_hull <- mod_score_env %>% 
  group_by(Fire, forest.type) %>% 
  slice(chull(NMDS1, NMDS2)) ## create convex hull by groups

p1 <- ggplot() +
  geom_point(aes(x = NMDS1, y = NMDS2, col = forest), data = mod_score_env) +
  geom_polygon(aes(x = NMDS1, y = NMDS2, fill = forest, color = forest), data = mod_score_hull, alpha = 0.3) +
  scale_fill_brewer(palette = "Paired") +
  scale_color_brewer(palette = "Paired")

p1
```

Just fire

```{r}
mod_score_hull2 <- mod_score_env %>% 
  group_by(Fire) %>% 
  slice(chull(NMDS1, NMDS2))

ggplot() +
  geom_point(aes(x = NMDS1, y = NMDS2, col = Fire), data = mod_score_env) +
  geom_polygon(aes(x = NMDS1, y = NMDS2, fill = Fire, color = Fire), data = mod_score_hull2, alpha = 0.3)
```


## facet by forest types and fire

```{r, , fig.height=5, fig.width=10}
p1 +
  facet_grid(~ forest.type)
```


```{r, , fig.height=5, fig.width=10}
p1 +
  facet_grid(~ Fire)
```


## simper analysis

```{r}
ss <- simper(comm.env.C, comm.env.E$Fire, permutations = 99)
summary(ss)
```

make new data.frame from the results
```{r}
ss_df <- data.frame(species = ss$before_after$species,
                    contribution = round(ss$before_after$average,3),
                    abun_before = round(ss$before_after$ava,3), 
                    abun_after = round(ss$before_after$avb,3),
                    p_value = ss$before_after$p)

ss_df %>% 
  filter(p_value <= 0.05)
```

